```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Activity monitoring exploratory data analysis

### Dataset Information

The data used here is available from: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

The variables included in this dataset are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

* date: The date on which the measurement was taken in YYYY-MM-DD format

* interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

### 1. Loading and Preprocessing the data

Let's load and preprocess the data!

```{r loadNpreprocess}
zipF <- "repdata_data_activity.zip"
outDir <- "activity"
unzip(zipfile = zipF, exdir = outDir)

activity <- read.csv("activity/activity.csv", header = TRUE, sep = ",")
activity$date <- as.Date(activity$date, "%Y-%m-%d")
dates <- unique(activity$date)
intervals <- unique(activity$interval)
```

### 2. Mean and Median total number of steps taken per day

Calculating number of steps taken each day

```{r steps}
stepsSplit <- split(activity$steps, activity$date)
totalStepsPerDay <- sapply(stepsSplit, sum, na.rm=TRUE)
```

Histogram of the total number of steps taken each day

```{r histogram}
plot(dates, totalStepsPerDay, main="Histogram of steps taken each day", xlab="Date", ylab="Frequency", type="h", lwd = 10)
```

Mean

```{r mean}
meanStepsPerDay <- sapply(stepsSplit, mean, na.rm=TRUE)
meanDf <- data.frame(dates, meanStepsPerDay, row.names=NULL)
meanDf
```

Median

```{r median}
medianStepsPerDay <- sapply(stepsSplit, median, na.rm=TRUE)
medianDf <- data.frame(dates, medianStepsPerDay, row.names=NULL)
medianDf
```

### 3. Average daily activity pattern

Time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

```{r timeseries}
intervalSplit <- split(activity$steps, activity$interval)
averageStepsPerInterval <- sapply(intervalSplit, mean, na.rm=TRUE)
plot(intervals, averageStepsPerInterval, type="l",
     main="Average number of steps per interval across all days", 
     xlab="Interval", ylab="Frequency")
```

Interval with maximum number of steps

```{r maxsteps}
maxIntervalDays <- max(averageStepsPerInterval, na.rm=TRUE)
maxIndex <- as.numeric(which(averageStepsPerInterval == maxIntervalDays))
paste("Interval with maximum number of steps: ", intervals[maxIndex])
```

### 4. Imputing missing values

Calculating the total number of missing values in the dataset 

```{r #NA}
noNAs <- sum(is.na(activity$steps))
paste("Total number of missing values: ", noNAs)
```

Replacing NA with mean total number of steps per day

```{r replaceNA}
meanSteps <- mean(meanStepsPerDay, na.rm = TRUE)
newActivity <- activity
newActivity$steps[which(is.na(activity$steps))] <- meanSteps
```

Histogram of the total number of steps taken each day with new datset

```{r newhist}
nstepsSplit <- split(newActivity$steps, newActivity$date)
ntotalStepsPerDay <- sapply(nstepsSplit, sum, na.rm=TRUE)
plot(dates, ntotalStepsPerDay, main="Histogram of steps taken each day", xlab="Date", ylab="Frequency", type="h", lwd = 10)
```

Mean and median total number of steps taken per day

New Mean

```{r newmean}
nmeanStepsPerDay <- sapply(nstepsSplit, mean, na.rm=TRUE)
nmeanDf <- data.frame(dates, nmeanStepsPerDay, row.names=NULL)
nmeanDf
```

New Median

```{r newmedian}
nmedianStepsPerDay <- sapply(nstepsSplit, median, na.rm=TRUE)
nmedianDf <- data.frame(dates, nmedianStepsPerDay, row.names=NULL)
nmedianDf
```

We can see that the only values that have changed are those days where all of the observations were missing. The rest of the observations are the same.

### 5.Differences in activity patterns between weekdays and weekends

Seperating days into weekends and weekdays

```{r}
newActivity$weekday <- weekdays(newActivity$date)
newActivity$DayType <- ifelse(newActivity$weekday=='Saturday' | newActivity$weekday=='Sunday', 'weekend','weekday')
```

Panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days. 

```{r}
StepsPerTimeDT <- aggregate(steps~interval+DayType,data=newActivity,FUN=mean)
library(ggplot2)
j <- ggplot(StepsPerTimeDT, aes(interval, steps))
j+geom_line(col="darkred")+ggtitle("Average steps per time interval: weekdays vs. weekends")+xlab("Time")+ylab("Steps")+theme(plot.title = element_text(face="bold", size=12))+facet_grid(DayType ~ .)
```
